---
title: "Advice sheet figures and tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advice sheet figures and tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.5, dpi = 72, fig.retina = 1
)
```

An advice sheet (`advice_en.qmd`, `advice_is.qmd` in each stock repository)
is built from a few objects of the `assessment_model` pipeline:
`assessment_hist`, `advice_hist`, `tac_hist`, `landings_by_gear`,
`landings_by_fishing_year_country`, and for category 3 stocks the rule's
components (`rfb_prognosis`, `chr_prognosis`). The `hr_advice_data_*()`
functions reshape them and the `hr_advice_plot_*()` / `hr_advice_table_*()`
functions draw the figures and tables. This vignette runs them on made-up
numbers for a toy stock.

```{r setup}
library(hafroreports)
library(patchwork)
```

## English and Icelandic

Every figure and table takes its labels from `getOption("hr.lang")` (`"en"`
or `"is"`, through `hr_label()` and the `hr_locale` table). Each sheet sets it
from its `lang:` front matter in `quarto_setup.R`:

```{r, eval = FALSE}
lang <- rmarkdown::metadata$lang
options(pax.lang = lang, hr.lang = lang)
```

```{r}
options(hr.lang = "en")
c(hr_label("biomass_index"), hr_label("recruitment_age", 3))
withr::with_options(list(hr.lang = "is"), c(hr_label("biomass_index"), hr_label("recruitment_age", 3)))
```

## Toy data

```{r}
set.seed(2)
assessment_year <- 2026
# Assessment summaries of the last five assessments (biomass in tonnes,
# recruitment in thousands), as assessment_hist
rec_all <- 80000 * exp(rnorm(length(1995:assessment_year), 0, 0.3))
assessment_hist <- do.call(rbind, lapply(2022:assessment_year, function(ay) {
  year <- 1995:ay
  ssb <- 60000 + 20000 * sin((year - 1995) / 6) + 1000 * (ay - 2024)
  hr <- ifelse(year == ay, NA, 0.25 + 0.1 * cos((year - 1995) / 5))
  rec <- rec_all[seq_along(year)] * (1 + 0.03 * (ay - 2024))
  out <- hr_assessment_template()[rep(1, length(year)), ]
  out$year <- year
  out$species <- 99L
  out$assessment_year <- as.integer(ay)
  out$median_SSB <- ssb
  out$low_SSB <- ssb * 0.8
  out$high_SSB <- ssb * 1.2
  out$median_refbio <- ssb * 1.6
  out$low_refbio <- ssb * 1.6 * 0.8
  out$high_refbio <- ssb * 1.6 * 1.2
  out$median_HR <- hr
  out$low_HR <- hr * 0.8
  out$high_HR <- hr * 1.2
  out$median_recruitment <- rec
  out$low_recruitment <- rec * 0.7
  out$high_recruitment <- rec * 1.3
  out$landings <- ifelse(year == ay, NA, hr * ssb * 1.6 / 1000)
  out
}))
# Reference points: biomass in thousand tonnes, as config.R
ref_points <- list(
  MGT_btrigger = 45, B_lim = 32, B_pa = 45,
  HR_mgt = 0.3, HR_msy = 0.3, HR_pa = 0.4
)
landings_by_gear <- tidyr::expand_grid(year = 1995:2025, gear_name = c("BMT", "LLN", "DSE", "Other")) |>
  dplyr::mutate(catch = runif(dplyr::n(), 1e6, 8e6)) # kg
```

## F- or HR-based stocks

`hr_advice_data_assessment()` turns the wide history into one row per year,
key and assessment, with English and Icelandic labels:

```{r}
data_assessment <- hr_advice_data_assessment(assessment_hist)
data_landings <- hr_advice_data_landings(landings_by_gear)
unique(data_assessment$key)
```

The four-panel stock development figure. In the sheets it is wrapped in
`ggiraph::girafe(ggobj = ...)` for the tooltips; here it is drawn as a static
plot:

```{r, fig.height = 6, fig.alt = "Four panels for the toy stock: catches by gear, recruitment, harvest rate with reference points and SSB with reference biomass."}
hr_advice_plot_landings(data_landings, assessment_year = assessment_year, year_start = 1995) +
  hr_advice_plot_recruitment(data_assessment, assessment_year, recruitment_age = 3) +
  hr_advice_plot_fpl(data_assessment, assessment_year, ref_points, fishing_pressure = "HR") +
  hr_advice_plot_ssb(data_assessment, assessment_year, ref_points)
```

`hr_advice_plot_fpl(fishing_pressure = "F")` draws F with `F_mgt`, `F_msy`,
`F_pa`. For F-based stocks the history has no reference biomass; series
without values are left out of `hr_advice_plot_ssb()`, but drop old empty
`refbio` rows first if the scale complains (`23-ple`).

The retrospective figure compares the last assessments:

```{r, fig.height = 5, fig.alt = "Retrospective panels of the toy stock: harvest rate, recruitment, reference biomass and SSB of the last five assessments."}
hr_advice_plot_retro(data_assessment, ref_points, assessment_year, fishing_pressure = "HR")
```

Axes are set from the data. A stock that needs a fixed range adds a scale to
the returned plot inside `suppressMessages()` (a new scale replaces the old
one, with a message).

## Tables

Reference points are kept in thousand tonnes for the figures;
`hr_advice_ref_table()` shows the biomass ones in tonnes
(`biomass_multiplier = 1000`). The basis table (in `config.R`) gives the row
labels and the basis of each point:

```{r}
ref_points_basis_table <- data.frame(
  render = c("MGT B~trigger~", "HR~MGT~", "B~lim~"),
  approach.en = c("Management plan", "Management plan", "Precautionary approach"),
  approach.is = c("Aflaregla", "Aflaregla", "Varúðarnálgun"),
  ref_point = c("MGT_btrigger", "HR_mgt", "B_lim"),
  basis.en = c("B~pa~", "Management plan", "B~loss~"),
  basis.is = c("B~pa~", "Aflaregla", "B~loss~")
)
hr_advice_ref_table(ref_points[c("MGT_btrigger", "HR_mgt", "B_lim")], ref_points_basis_table)
```

The TAC table joins `advice_hist`, `tac_hist` and the landings by fishing
year and country (kg):

```{r}
advice_hist <- data.frame(
  assessment_year = 2022:2026,
  advice_period = paste(2022:2026, 2023:2027, sep = "/"),
  advice = c(28000, 30500, 29800, 31200, 32900)
)
tac_hist <- data.frame(assessment_year = 2022:2026, ices_area = "5a", tac = advice_hist$advice)
landings_by_fishing_year_country <- data.frame(
  fishing_year = rep(paste(2022:2025, 2023:2026, sep = "/"), each = 2),
  country = rep(c("Iceland", "Other"), 4),
  catch = c(27e6, 0.5e6, 29e6, 0.6e6, 30e6, 0.4e6, 30.5e6, 0.3e6)
)
data_tac <- hr_advice_data_tac(advice_hist, tac_hist, landings_by_fishing_year_country)
hr_advice_table_tac(data_tac, foreign_footnote = FALSE)
```

Stock-specific notes on rows go in `footnotes` (in `config.R`, e.g.
`tac_table_footnotes` in `02-had-targets`), not in the package.

## Category 3 stocks

Index-based stocks (`25-wit`, `13-cas`, `14-mon`, `24-lem`, `26-meg`,
`27-dab`, `60-norway-redfish`; `04-whg` uses the chr rule) keep their
indicators in the same history columns: the biomass index as `SSB`, a
juvenile index as `recruitment`, the fishing pressure proxy L~F=M~ / L~mean~
as `F`. Relabel them with `labels`:

```{r, fig.alt = "Toy biomass index with I_trigger and the index A and B means, and the fishing pressure proxy with F_MSY proxy."}
index_hist <- do.call(rbind, lapply(2025:2026, function(ay) {
  year <- 1996:ay
  b <- 2500 + 800 * sin((year - 1996) / 4) + 30 * (year - 1996)
  out <- hr_assessment_template()[rep(1, length(year)), ]
  out$year <- year
  out$species <- 99L
  out$assessment_year <- as.integer(ay)
  out$median_SSB <- b
  out$low_SSB <- b * exp(-1.96 * 0.2)
  out$high_SSB <- b * exp(1.96 * 0.2)
  out$median_F <- ifelse(year < ay, 1 + 0.1 * cos(year / 3), NA)
  out
}))
data_index <- hr_advice_data_assessment(
  index_hist,
  labels = list(
    en = c(SSB = "Biomass index", F = "Fishing pressure proxy"),
    is = c(SSB = "Lífmassavísitala", F = "Vísitala veiðihlutfalls")
  )
)
ref_index <- list(I_trigger = 1900, F_msy_proxy = 1)
hr_advice_plot_index(data_index, assessment_year, ref_index, index_ab = TRUE, index_ab_span = "periods") +
  hr_advice_plot_fproxy(data_index, assessment_year, ref_index, y_pad = 0.1, points = TRUE)
```

`index_ab = TRUE` draws index A (mean of the last two years) and index B (the
three years before) of the rfb rule.

The advice calculation tables take the rule's components as a data frame
with `component` and `value` (from `dlsrules::rfb_rule()` or
`dlsrules::fproxy_rule()`; component names in any case). The row layout and
the English and Icelandic text are in `inst/extdata/rfb_prognosis_base.csv`
and `chr_prognosis_base.csv`:

```{r}
rfb_prognosis <- data.frame(
  component = c(
    "previous_catch_advice", "index_A", "index_B", "index_ratio",
    "mean_catch_length", "msy_proxy_length", "fpp_ratio_inv",
    "last_index_value", "Itrigger", "biomass_safeguard", "multiplier",
    "initial_catch_advice", "stability_clause_applied", "catch_advice",
    "advice_change"
  ),
  value = c(1200, 3350, 3100, 1.081, 41.5, 37.5, 0.904, 3400, 1900, 1, 0.95, 1114, 0, 1114, -7.2)
)
hr_advice_table_rfb(rfb_prognosis, assessment_year, biannual = TRUE)
```

The same table in Icelandic:

```{r}
withr::with_options(list(hr.lang = "is"), hr_advice_table_rfb(rfb_prognosis, assessment_year))
```

`hr_advice_table_chr()` works the same way for the constant harvest rate rule
(`04-whg`).

## Things to know

- The banner (`hr_advice_banner()`) and the text read `tac` and
  `tac_last_year` from `advice_hist`/`tac_hist`; they must hold the published
  rows (see `vignette("history-tables")`).
- Category 3 advice is often biennial: `biannual = TRUE` in the rfb table, two
  rows in `advice_hist`, and a banner `note` for the second year.
- Fixed haddock values used to be built into these functions (axis limits,
  recruitment age, TAC footnotes, retro colours). They are arguments now;
  when a figure looks wrong for a stock, look for the argument before writing
  a stock copy of the function.

## AI use

This vignette was drafted with Claude (Anthropic) in October 2026 and has not
yet been checked by a person (MFRI policy on AI use).
