---
title: "Input data for SAM and Muppet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Input data for SAM and Muppet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

The age-based assessments (SAM or Muppet) take one table, `input_data`, with a
row per year and age: catch in numbers and catch weights, the survey indices,
stock weights, maturity and natural mortality. In a stock repository it is
built by a chain of targets in `script_assessment_model.R`:

| Target | Function | Gives |
|---|---|---|
| `pax_db` | `pax::pax_from_mar()` | the local DuckDB with stations, length and age samples, landings |
| `input_data_lw_pred` | `hr_input_data_lw()` | weight at length (GAM fit to the survey) |
| `input_data_maturity_key` | `hr_input_data_maturity_key()` | maturity at length (and age) |
| `input_data_igfs_index`, `input_data_agfs_index` | `hr_input_data_si_index()` | survey abundance, weight and maturity at age |
| `input_data_comm_index` | `hr_input_data_si_index()` | catch at age (commercial samples raised to landings) |
| `input_data_landings` | `hr_input_data_landings()` | landings by year |
| `input_data` | `hr_input_data_combine()` | the model input table |
| `sam_dat` | `hr_sam_dat()` | the SAM data object |

`02-had-targets` is the template; `03-sai`, `23-ple`, `08-usk` and `01-cod` show
how the arguments differ between stocks. This vignette runs the chain on a
small simulated database, so it builds without the MFRI database.

```{r setup}
library(hafroreports)
```

## The pax database

In a stock repository, `pax_db` is built from the MFRI database (Oracle,
through the `mar` package) once, or opened from a copy:

```{r, eval = FALSE}
tar_target(
  pax_db,
  if (nzchar(Sys.getenv("PAX_SOURCE_DB"))) {
    pax_connect(Sys.getenv("PAX_SOURCE_DB"))
  } else {
    pax_from_mar(species, year_start, year_end)
  },
  format = pax_tar_format_duckdb()
)
```

Here we build a toy database with the same table names (`station`, `ldist`,
`aldist`, `measurement`, `landings`, `lw_coeffs`) from simulated fish. The
code is long only because it simulates the samples; the point is that every
`hr_input_data_*()` function takes a pax connection and works on these tables.

```{r child = "_toy-pax-db.Rmd"}
```

```{r}
dplyr::tbl(pax_db, "station") |> dplyr::count(sampling_type) |> dplyr::collect()
```

## Weight and maturity keys

The weight-at-length key is a GAM fitted to the survey otoliths:

```{r}
lw <- hr_input_data_lw(pax_db, sampling_type = 30, prediction_length_range = 1:120)
lw[lw$length %in% c(20, 40, 60), ]
```

The maturity key fits a logistic model of maturity at length to the survey
otoliths. Measured maturity at age is kept, with the model predictions marked
by `age = NA`:

```{r}
mat_key <- hr_input_data_maturity_key(pax_db, lgroups = seq(0, 120, 5))
mat_key |> dplyr::filter(is.na(age), lgroup %in% c(30, 45, 60))
```

The options the stocks use:

- `by_year = TRUE` fits a length + year model (`03-sai`, `23-ple`), with
  `copy_years` for years without data.
- `regions` with more than one region fits `log(lgroup) * region`; with one
  region the formula has no region term (earlier versions failed with
  "contrasts can be applied only to factors with 2 or more levels").
- `sex = 2`, `mature_above`, `immature_below` reproduce the plaice key
  (`23-ple`).
- `aged_only = FALSE` also uses otoliths with a maturity stage but no age.
  pax relabels otoliths without an otolith number as `LEN`, so check where
  the maturity samples are in your stock's data.

## Survey indices

`hr_input_data_si_index()` raises the length distributions of each tow to the
tow area, scales them to the strata, splits them into ages with an age-length
key and sums abundance (`n`, thousands), mean weight (`mw`, g) and maturity
(`mat`) by year and age.

Strata can be assigned from the tow position (`strata_name` alone,
`pax::pax_si_scale_by_strata()`), or from a fixed station list
(`strata_stations`). The old tidypax indices used the station list
(`biota.strata_stations`, in `pax_db` as the `strata_stations` table), and
assigning by position changed the saithe spring survey index by -47% to +64%
in single years. Use the station list to reproduce a published index. Here the
toy station list carries the stratum areas itself:

```{r}
smb <- hr_input_data_si_index(
  pax_db,
  sampling_type = 30,
  lw_key = lw,
  maturity_key = mat_key,
  regions = list(all = 101:115),
  strata_stations = toy_strata_stations
)
smb |> dplyr::filter(year == 2024) |> dplyr::collect() |> head(4)
```

In a stock repository the station list comes from `pax_db` and the areas from
the pax strata table:

```{r, eval = FALSE}
hr_input_data_si_index(
  pax_db,
  sampling_type = 30,
  tow_number = 0:35,
  lw_key = input_data_lw_pred,
  maturity_key = input_data_maturity_key,
  strata_name = "old_strata",
  strata_stations = strata_stations |>
    dplyr::filter(sampling_type == 30, stratification == "old_strata"),
  regions = list(all = 101:115)
)
```

Other survey options: `gear_id_filter` and `tow_number` select the index
tows; `haul_scalar` down-weights named very large hauls (`03-sai`); `ygroup`
pools years with few otoliths in the age-length key (`23-ple`), and `alk` or
`key_year` give a key the function can't build. Use `hr_pool_years()` rather
than pax's own `ygroup`, which fails in DuckDB.

## Catch at age

The same function gives the catch at age from the commercial samples: an
age-length key by gear group and half year, raised to the landings with
`scale_by_landings = TRUE`.

Don't give a `tow_number` filter for commercial samples. For them it is the
haul number, which can exceed 35; the template's `tow_number = 0:35` dropped
3–14% of the saithe and haddock samples in recent years and shifted the catch
at age. In the toy data:

```{r}
dplyr::tbl(pax_db, "station") |>
  dplyr::filter(sampling_type == 1) |>
  dplyr::summarise(samples = dplyr::n(), haul_above_35 = sum(ifelse(tow_number > 35, 1, 0))) |>
  dplyr::collect()
```

```{r}
comm <- hr_input_data_si_index(
  pax_db,
  sampling_type = 1,
  lw_key = lw,
  tgroup = list(t1 = 1:6, t2 = 7:12),
  gear_group = list(BMT = "BMT", LLN = "LLN"),
  scale_by_landings = TRUE
)
comm |> dplyr::filter(year == 2024) |> dplyr::collect() |> head(4)
```

Missing values in the samples or the landings are dropped silently unless
you say what they are. The function reports how many samples it leaves out:

- `gridcell_na`: samples without a position match no region (plaice lost the
  whole of 1990 without `gridcell_na = 2741`).
- `sample_gear_na`, `landings_gear_na`, `landings_month_na`: what unknown
  gears and months were in the old scripts (saithe: DSE for samples, BMT and
  month 6 for landings).
- `pax::pax_add_other()` in `gear_group` gives a default group, for gears
  outside the listed groups.
- `sam_use_10_11_first_2_years = TRUE` adds survey otoliths to the commercial
  key in the first two years. It is a haddock choice; saithe and plaice don't
  use it.

## Combine

`hr_input_data_combine()` joins the indices and landings, fills gaps and
scales the catch so that catch × catch weight equals the landings each year.
With the defaults it only fills weights with the mean for the age; the
stock's own rules are arguments (maturity cut-offs, values before the survey
started, running means, M), as in `03-sai`:

```{r, eval = FALSE}
hr_input_data_combine(
  year_start, year_end,
  age_start = 0,
  input_data_comm_index = input_data_comm_index,
  input_data_igfs_index = input_data_igfs_index,
  input_data_agfs_index = input_data_agfs_index,
  input_data_landings = input_data_landings,
  M = 0.2,
  immature_below_age = 4,
  mature_above_age = 9,
  maturity_fixed = data.frame(age = 4:9, maturity = c(0.08, 0.185, 0.371, 0.61, 0.79, 0.912), year_before = 1985),
  maturity_smooth_years = 5,
  weight_fill_year = 1985,
  stock_weight_smooth_above = 8,
  stock_weight_smooth_years = 4
)
```

The toy stock has no autumn survey, so we pass an empty table:

```{r}
input_data <- hr_input_data_combine(
  year_start = 2015,
  year_end = 2024,
  age_start = 1,
  input_data_comm_index = comm,
  input_data_igfs_index = smb,
  input_data_agfs_index = data.frame(year = numeric(), age = numeric(), n = numeric()),
  input_data_landings = hr_input_data_landings(pax_db),
  immature_below_age = 2,
  mature_above_age = 7
)
input_data |> dplyr::filter(year == 2024) |> head(5)
```

Check that catch × catch weight gives the landings (kg):

```{r}
input_data |>
  dplyr::group_by(year) |>
  dplyr::summarise(catch_kg = sum(catch * catch_weight, na.rm = TRUE)) |>
  dplyr::left_join(dplyr::collect(hr_input_data_landings(pax_db)), by = "year") |>
  head(3)
```

Leave `age_end = NULL` (all ages in the data) and let the model form the plus
group. Cutting ages here drops the old fish, and scaling to landings then
moves their weight onto the younger ages: saithe SSB was about 30% too low.

## SAM

`hr_sam_dat()` turns `input_data` into a SAM data object, one `hr_sam_*()`
helper per matrix. The defaults are the spring survey only and no F or M
before spawning; give the stock's values explicitly (haddock: `pf = 0.4`,
`pm = 0.3`). The wrong proportions put saithe SSB 14% low while every input
matched.

```{r, eval = requireNamespace("SAMutils", quietly = TRUE) && requireNamespace("stockassessment", quietly = TRUE)}
model_dat <- input_data |> dplyr::filter(year <= 2023)
sam_dat <- hr_sam_dat(model_dat = model_dat, minage = 1, maxage = 10)
names(sam_dat)[1:5]
```

The configuration (`hr_sam_conf()`) is the haddock one; each stock keeps its
own `*_sam_conf()` in `R/sam.R`, written from the old `model_setup.R`. The
fit itself is `SAMutils::full_sam_fit(sam_dat, sam_conf)`. The SAM chunks
above run only when SAMutils and stockassessment are installed.

## Muppet

`hr_muppet_input_datafiles()` writes the same table as the Muppet data files
(`01-cod`). `haddock_fills = FALSE` writes missing weights and maturity as
missing (-1) instead of haddock values:

```{r}
files <- hr_muppet_input_datafiles(
  input_data,
  year_start = 2015,
  year_end = 2024,
  age_end = 10,
  haddock_fills = FALSE,
  smb_year_start = 2015,
  smh_year_start = 2015,
  survey_ages = 1:10
)
names(files)
cat(head(strsplit(files[["Files/catchandstockdata.dat"]], "\n")[[1]], 4), sep = "\n")
```

```{r, include = FALSE}
DBI::dbDisconnect(pax_db)
```

## AI use

This vignette was drafted with Claude (Anthropic) in October 2026 and has not
yet been checked by a person (MFRI policy on AI use).
