---
title: "Tech report figures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tech report figures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4, dpi = 72, fig.retina = 1
)
```

The technical reports (`techreport_en.qmd`, `techreport_is.qmd`) document
the data and the model behind the advice. The `hr_techreport_*()` functions
draw the standard figures. They take one of three kinds of input:

- **the pax database** (`pax_db`): landings, sampling, length distributions,
  survey maps, e.g. `hr_techreport_plot_landings_gear()`,
  `hr_techreport_plot_survey_ldist()`, `hr_techreport_plot_sampling_overview()`;
- **the model input table** (`input_data`, see `vignette("input-data")`):
  catch and survey at age, weights, maturity, e.g.
  `hr_techreport_plot_catch_age_bubble()`, `hr_techreport_plot_maturity()`,
  `hr_techreport_plot_wm_left()`;
- **model or rule results**: the SAM fit (`hr_techreport_plot_samres_resid()`,
  needs SAMutils), the Muppet fit, or for category 3 stocks the rfb rule
  (`hr_techreport_plot_rfb_*()`).

This vignette draws a few of each on the simulated toy database of
`vignette("input-data")` (the code is hidden here) and on made-up rule
results. Labels follow `getOption("hr.lang")`, as in the advice sheets.

```{r setup}
library(hafroreports)
options(hr.lang = "en")
```

```{r child = "_toy-pax-db.Rmd", echo = FALSE}
```

```{r}
# The toy input_data, as in vignette("input-data")
lw <- hr_input_data_lw(pax_db, sampling_type = 30, prediction_length_range = 1:120)
smb <- hr_input_data_si_index(
  pax_db, sampling_type = 30, lw_key = lw,
  maturity_key = hr_input_data_maturity_key(pax_db, lgroups = seq(0, 120, 5)),
  strata_stations = toy_strata_stations
)
comm <- hr_input_data_si_index(
  pax_db, sampling_type = 1, lw_key = lw, tgroup = list(t1 = 1:6, t2 = 7:12),
  gear_group = list(BMT = "BMT", LLN = "LLN"), scale_by_landings = TRUE
)
input_data <- hr_input_data_combine(
  2015, 2024, age_start = 1,
  input_data_comm_index = comm, input_data_igfs_index = smb,
  input_data_agfs_index = data.frame(year = numeric(), age = numeric(), n = numeric()),
  input_data_landings = hr_input_data_landings(pax_db)
)
```

## From the pax database

Landings by gear group:

```{r}
hr_techreport_plot_landings_gear(pax_db, gear_group = list(BMT = "BMT", LLN = "LLN"))
```

Survey length distributions by year:

```{r, fig.height = 6}
hr_techreport_plot_survey_ldist(pax_db, sampling_type = 30)
```

The pax `ldist` table is already raised to the counted fish when it is
imported (`pax_mar_ldist()`). The hafroreports survey length and location
figures use it as it is; in your own code, don't apply
`pax::pax_ldist_scale_abund()` again, but pass
`ldist_tbl = dplyr::tbl(pax_db, "ldist") |> pax::pax_ldist_scale_round()` to
`pax::pax_ldist_by_year()` (`25-wit`, `04-whg`). Raising twice made whiting
length distributions a few percent high in single years.

Several figures depend on stock-specific settings: `gear_group` and
`depth_class` (use the stock's gear groups and depth classes), `regions` and
`keep_order` for the regional figures, `strata_stations` for the survey by
area. `hr_techreport_plot_catchdepth()` returns a two-panel patchwork; change
its fill scale with `&`, not `+` (`23-ple`).

## From input_data

Catch at age as bubbles (area proportional to numbers):

```{r}
hr_techreport_plot_catch_age_bubble(input_data, year_start = 2015, age_start = 2)
```

Maturity at age by year:

```{r}
hr_techreport_plot_maturity(input_data, year_start = 2015)
```

## Category 3: the rfb rule

The rfb figures take the survey index of the rule (columns `index`, `year`,
`b` in tonnes and `b_cv`), the catch length distributions (`year`, `length`,
`n`), the rule's reference points and the rule's components (as in
`vignette("advice-sheets")`):

```{r}
survey_index <- data.frame(
  index = "smb_harv",
  year = 1996:2026,
  b = 2500 + 800 * sin((1996:2026 - 1996) / 4) + 30 * (0:30),
  b_cv = 0.2
)
ldist <- expand.grid(year = 2005:2025, length = 10:70)
ldist$n <- dnorm(ldist$length, 40 + (ldist$year - 2005) / 10, 8) * 1000
ref_points <- list(
  I_trigger = 1900, I_lim = 1360, Lc = 30, Linf = 70,
  target_reference_length = 0.75 * 30 + 0.25 * 70, F_msy_proxy = 1
)
rfb_prognosis <- data.frame(
  component = c("index_A", "index_B", "mean_catch_length"),
  value = c(3350, 3100, 45.5)
)
```

```{r, fig.height = 6}
library(patchwork)
(hr_techreport_plot_rfb_index(survey_index, "smb_harv", rfb_prognosis, ref_points, assessment_year = 2026) +
  hr_techreport_plot_rfb_lc(ldist, ref_points)) /
  (hr_techreport_plot_rfb_ml(ldist, ref_points, data_year = 2025, rfb_prognosis = rfb_prognosis) +
    hr_techreport_plot_rfb_f(ldist, ref_points, year_start = 2005)) &
  hr_techreport_rfb_theme()
```

`hr_techreport_dsl_basis()`, `hr_techreport_plot_dsl_lc()` and
`hr_techreport_plot_dsl_ml()` draw the length-based figures of stocks whose
rule uses survey length data as well.

## Model fits

The SAM figures take `SAMutils` results and only run where SAMutils is
installed:

```{r, eval = FALSE}
hr_techreport_plot_samres_resid(sam_fit)
hr_techreport_plot_samres_combfit(sam_fit, input_data)
```

Copy a figure into the stock's `R/techreport.R` only when an argument can't
do what it needs (e.g. an age-based reference biomass in the SAM summary,
`03-sai`), and note why.

```{r, include = FALSE}
DBI::dbDisconnect(pax_db)
```

## AI use

This vignette was drafted with Claude (Anthropic) in October 2026 and has not
yet been checked by a person (MFRI policy on AI use).
